---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


### Loading and preprocessing the data

I assume that the file `activity.cgesv` is placed in the same folder as this file (`PA1_template.Rmd`).

```{r}
data <- read.csv("activity.csv")
data$date <- as.Date(data$date , format = "%Y-%m-%d")
```

### What is mean total number of steps taken per day?

Total number of steps per day:

```{r}
steps.per.day <- aggregate(data$steps, by = list(data$date), sum)
colnames(steps.per.day) <- c("Date", "Steps")
summary(steps.per.day)
```

```{r}
hist(steps.per.day$Steps, breaks = 20, main = "Total number of steps taken each day", xlab = "Steps")
```

Mean number of steps taken per day (over the days for which data is present):

```{r}
mean(steps.per.day$Steps, na.rm = TRUE)
```

Median number of steps taken per day (over the days for which data is present):

```{r}
median(steps.per.day$Steps, na.rm = TRUE)
```


### What is the average daily activity pattern?

Time series plot of average number of steps taken over each 5-minute interval, averaged across all days

```{r}
steps.per.interval <- aggregate(data$steps, by = list(data$interval), mean, na.rm = TRUE)
colnames(steps.per.interval) <- c("Interval", "Steps")
plot(steps.per.interval$Interval, steps.per.interval$Steps, type="l", main = "Average number of steps over 5-minute interval", 
     xlab = "Interval", ylab = "Steps")
```

The 5-minute interval that contains the maximum number of steps (on average across all days) is:

```{r}
steps.per.interval[which.max(steps.per.interval$Steps), ]
```

### Imputing missing values

Total number of missing values in the dataset:

```{r}
sum(is.na(data$steps))
```

To fill the missing values, I'll use the median for that 5-minute interval based on data from the rest of the days. The dataset equal to the original but with the missing data filled in is:

```{r}
median.steps.per.interval <- aggregate(data$steps, by = list(data$interval), median, na.rm = TRUE)
colnames(median.steps.per.interval) <- c("interval", "median.steps")

imputed.data <- merge(data, median.steps.per.interval, sort = FALSE)
missing <- is.na(imputed.data$steps)
imputed.data$steps[missing] <- imputed.data$median.steps[missing]
imputed.data$median.steps <- NULL
```

To compare resulting dataset to the original one, we repeat the steps done in the first part of the assignment: build a histogram of number of steps per day...

```{r}
imputed.steps.per.day <- aggregate(imputed.data$steps, by = list(imputed.data$date), sum)
colnames(imputed.steps.per.day) <- c("Date", "Steps")
hist(imputed.steps.per.day$Steps, breaks = 20, main = "Total number of steps taken each day", xlab = "Steps")
```

...and calculate mean and median values for them:

```{r}
mean(imputed.steps.per.day$Steps)
median(imputed.steps.per.day$Steps)
```

Both mean and median are lower than for data set with missing values, because imputed values are mostly smaller than average (second bucket of the histogram).

### Are there differences in activity patterns between weekdays and weekends?

Add a new factor to the dataset with "weekday" and "weekend" levels.

```{r}
week.data <- imputed.data
week.data$weekday <- factor(weekdays(week.data$date) %in% c("Saturday", "Sunday"), levels = c(FALSE, TRUE), labels = c("weekday", "weekend"))
summary(week.data)
```

Time series plot of average number of steps taken over each 5-minute interval, averaged across all weekdays and all weekends separately:

```{r}
avg.steps <- aggregate(week.data$steps, by = list(week.data$interval, week.data$weekday), mean)
colnames(avg.steps) <- c("Interval", "Day", "Steps")

par(mfrow = c(2, 1))
plot(avg.steps[avg.steps$Day == "weekday", 1], avg.steps[avg.steps$Day == "weekday", 3], 
     type="l", main = "Weekdays", 
     xlab = "Interval", ylab = "Steps")
plot(avg.steps[avg.steps$Day == "weekend", 1], avg.steps[avg.steps$Day == "weekend", 3], 
     type="l", main = "Weekends", 
     xlab = "Interval", ylab = "Steps")
```

We can see that on the weekends the movement starts later and ends later.
